Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-05-07 paper/tidy-missing-data-paper.Rmd
completed section 4

Common missing data operations can be considered verbs, in the tidyverse sense. For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**. Data can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with `NA`. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. Data matrix can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`. 
The function is called  `miss_scan_count()`, and it returns a table of occurrences of that value for each variable. Two more data objects, `common_na_numbers` and `common_na_strings` provide a list of common NA values for numbers and characters, to help users check for the typical representations of missing.

@@

### replace: Replacing missing value label with another {#verbs-replace-with}

Once the different representations of possible missing values have been identified, these values can be replaced using the function, `replace_with_na()`.  A dataset could have the values `-99` in the `x` column replaced with the following code:
`replace_with_na(dat_ms, replace = list(x = -99))`. For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` makes a conditional change on columns **if** they meet some condition (such as `is.numeric` or `is.character`).

@@

Summary information such as the number of missings in a given case can be useful in understanding missingness structure, and even more so when this information is kept alongside the data. Functions starting with `add_` exist in `dplyr` to add count information for particular groups or conditions to the data. `naniar` includes functions to add missingness summary information to the data, such as the number or proportion of missingness, the missingness cluster, or if there are any missings, using: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values or nice labels for any missing values with `add_label_shadow()` and `add_label_missings()`. An example use of these features is in @Tierney2015, where the proportion of
missings is used as the outcome in a model, to identify variables and values
important in predicting missingness structures.

@@

_Nabular_ data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. _Nabular_ data can be created with `nabular()`, as below:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. These special types of missingness can be encoded in the shadow matrix of  _nabular_  data, using `recode_shadow()`. Using `recode_shadow()` requires specifying the variable you want to contain the flagged value, the condition you want this to occur at, and a suffix for the new type of missing value. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable x:

@@

`naniar` does not reinvent the wheel for imputation, as there are many robust R packages for imputation \@ref(imputation). Instead, _nabular_ data is designed to work with existing imeplementation. `naniar` does provide a few imputation methods to facilitate exploration and visualization: `impute_mean`, `impute_median`, and `impute_below`. While useful to explore structure in missingness, they are not recommended for use in analysis. `impute_below` imputes values below the minimum value, with some jitter (small random noise) to reduce overplotting. It has options to change the default amount of shift, and jitter.

@@

To evaluate imputations they need to be tracked, which is achieved by combining the verbs `nabular`, `impute_`, and `add_label_shadow`. The missing values can then be referred to by their shadow variable, `_NA`. The missingness of any observation can be referred to with `any_missing` (Figure \@ref(fig:track-impute-example)). The code chunk below shows the track pattern, first using `nabular`, then imputing with `impute_lm`, and adding a `label_shadow`:

@@

Missing values can be mapped to a graphical element in a scatterplot by setting the `color` aesthetic in ggplot to `any_missing` (Figure \@ref(fig:track-impute-example)A), or in a density plot using the `fill = any_missing`, (Figures \@ref(fig:track-impute-example)B) and \@ref(fig:track-impute-example)C). Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows (shown below). This shows mean and median values are similar, but the minimum and maximum values are different.

@@

```{r impute-summary, echo = TRUE}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max)) 
```

>>>>

Common missing data operations can be considered verbs, in the tidyverse sense. For missing data, these include: **scan**, **replace**, **add**, **shadow**, **impute**, **track**, and **flag**. Data can be **scanned** to find possible missings not coded as `NA`. These values can then be **replaced** with `NA`. To facilitate exploration, summaries of missingness can be **added** as a summary column to the original data. The data can be augmented with the **shadow matrix** values, helping explore missing data, as well as facilitating the process of **imputing**, and **tracking**. Finally, unusual or specially coded missing values can be **flagged**.

@@

This operation is used to search the data for specific conventional representations of missings, such as, "N/A", "MISSING", `-99`. 
This is implemented in the function `miss_scan_count()`, which returns a table of occurrences of that value for each variable. A list of common NA values for numbers and characters, can be provided to help check for typical representations of missings. These are implemented in `naniar` as `common_na_numbers` and `common_na_strings`, for numbers and characters.

@@

### replace: Replacing values with missing values {#verbs-replace-with}

Once possible missing values have been identified, these values can be replaced. For example, a dataset could have the values `-99` meaning a missing value. `naniar` implements replacement with the function, `replace_with_na()`. Values -99 could be replaced in the `x` column with:
`replace_with_na(dat_ms, replace = list(x = -99))`. For operating on multiple variables, there are scoped variants for `replace_with_na`: `_all`, `_if`, and `_at`. This means `replace_with_na_all` operates on **all** columns, `replace_with_na_at` operates **at** specific columns, and `replace_with_na_if` makes a conditional change on columns **if** they meet some condition (such as `is.numeric` or `is.character`).

@@

Understanding the missingness structure can be improved by adding summary information alongside the data. For example, in @Tierney2015, the proportion of
missings in a row is used as the outcome in a model to identify variables
important in predicting missingness structures. `naniar` implements a series of functions to add these missingness summaries to the data, starting with `add_`. These are inspired by the `add_count` function in `dplyr`, which adds count information for specified groups or conditions. `naniar` provides operations to add the number or proportion of missingness, the missingness cluster, or the precense of any missings, with: `add_n_miss()`, `add_prop_miss()`, `add_miss_cluster`, and `add_any_miss()`, respectively (Table \@ref(tab:add-missing-info)). There are also functions for adding information about shadow values, and readable labels for any missing values with `add_label_shadow()` and `add_label_missings()`.

@@

_Nabular_ data has the shadow matrix column-bound to existing data. This facilitates visualization and summaries, and allows for imputed values to be tracked. _Nabular_ data can be created with `nabular()`:

@@

Unusual or spurious data values are often identified and `flagged`. For example, there might be special codes to mark an individual dropping out of a study, known instrument failure in weather instruments, or for values censored in analysis. `naniar` provides tools to encode these special types of missingness in the shadow matrix of  _nabular_  data, using `recode_shadow()`. This requires specifying the variable to contain the flagged value, the condition for flagging, and a suffix. This is then recoded as a new factor level in the shadow matrix, so every column is aware of all possible new values of missingness. For example, the values -99 and -98 could be recoded to mean a broken machine sensor for the variable `x` with the following:

@@

`naniar` does not reinvent the wheel for imputation, since instead working with existing methods. However, `naniar` provides a few imputation methods to facilitate exploration and visualization: `impute_below`, `impute_mean`, and `impute_median`. While useful to explore structure in missingness, they are not recommended for use in analysis. `impute_below` imputes values below the minimum value, with some controlable jitter (random noise) to reduce overplotting.

@@

To evaluate imputations they need to be tracked. This is achieved by first using `nabular` data, then imputing, and imputed values can then be referred to by their shadow variable, `_NA` \@ref(fig:track-impute-example)). The code below shows the track pattern, first using `nabular`, then imputing with `impute_lm`. A label is added to facilitate identifying missings with `label_shadow`:

@@

Multiple missing or imputed values can be mapped to a graphical element in ggplot2 by setting the `color` or `fill` aesthetic in ggplot to `any_missing`, a result of the `add_label_shadow()` function. (Figure \@ref(fig:track-impute-example)A - C). Imputed values can also be compared to complete case data, grouping by `any_missing`, and then summarizing, similar to other dplyr summary workflows shown below in Table \@ref(tab:impute-summary-out), showing similarities and differences in imputation methods.

@@

```{r impute-summary, echo = TRUE, eval = FALSE}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max)) 
```

```{r impute-summary-out, echo = FALSE}
aq_imputed %>%
  group_by(any_missing) %>%
  summarise_at(.vars = vars(Ozone),
               .funs = lst(min, mean, median, max))  %>% 
  kable(caption = "Summarising values of imputed vs non imputed values. Comparing imputed values (denoted as 'Missing', since they were previously missing), the mean and median values are similar, but the minimum and maximum values are very different.")
```
